Report on the Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4)

### 2017-04-28 WSSSPE4_report/wg_codemeta.tex
tense and person changes

The primary objective of this working group was to find ways to help the CodeMeta project come to fruition. CodeMeta seeks in part to create a ``Rosetta Stone'' for software metadata to facilitate retaining such metadata between repositories, services, registries, indexers, publishers, citation managers, and other entities that create, ingest, use, and/or store metadata about software. The project also wants to establish a JSON-LD schema as a tool for making metadata machine-readable~\cite{CodeMeta_schema}.

@@

Our plans are for the working group to engage with those already working on CodeMeta, to examine the existing CodeMeta crosswalk table to see what improvements and additions might be made, and to determine how to engage the community and provide ongoing social engagement and structure. Further, the group wants to assist in the implementation of the specifications and, by providing an outsider's view, contribute suggestions for more understandable project documentation. Finally, the WG seeks a better way or ways to present the crosswalk table to make it more easily understood and consumable by research software communities.

@@

With a comparatively large working group, the use of the established CodeMeta Github repository's issue tracking, and the quick responsiveness of CodeMeta lead Matt Jones to a barrage of questions, comments, and logged issues, several of our SMART steps were completed at WSSSPE; these are identified below in italics.

>>>>

The primary objective of this working group is to find ways to help the CodeMeta project come to fruition. CodeMeta seeks in part to create a ``Rosetta Stone'' for software metadata to facilitate retaining such metadata between repositories, services, registries, indexers, publishers, citation managers, and other entities that create, ingest, use, and/or store metadata about software. The project also wants to establish a JSON-LD schema as a tool for making metadata machine-readable~\cite{CodeMeta_schema}.

@@

The working group plans to engage with those already working on CodeMeta, to examine the existing CodeMeta crosswalk table to see what improvements and additions might be made, and to determine how to engage the community and provide ongoing social engagement and structure. Further, the group wants to assist in the implementation of the specifications and, by providing an outsider's view, contribute suggestions for more understandable project documentation. Finally, the WG seeks a better way or ways to present the crosswalk table to make it more easily understood and consumable by research software communities.

@@

With a comparatively large working group, the use of the established CodeMeta Github repository's issue tracking, and the quick responsiveness of CodeMeta lead Matt Jones to a barrage of questions, comments, and logged issues, several of the group's SMART steps were completed at WSSSPE; these are identified below in italics.

### 2017-04-28 WSSSPE4_report/wg_letters.tex
tense and person changes

We believe that the authors of scientific software provide important
services to departments that are no less than strictly disciplinary
research. Consequently, in order to level the playing field with more
disciplinary candidates for hires, tenure, and promotion requires that
letter writers be aware of how their letters will be read by
committees. It also requires that committees be aware that such
letters often look different and may provide a different perspective
of how a candidate's achievements should be assessed. For example, in
mathematics a typical candidate would be evaluated on the difficulty
and depth of the statements she may have proven in their papers. On
the other hand, an author of mathematical software would likely be
evaluated based on the impact of her software, or the number of citations of
the publications that describe it. She may also be evaluated by how
widely the software is used \textit{outside} mathematics, a criterion
that is rarely used for more disciplinary mathematicians.

@@

Affecting letter writers and readers essentially requires raising
awareness of the issue. Various organizations have attempted to do so
through letters to the community, adjusting definitions of what counts
in science, etc. Specifically, we are aware of the following
resources:

@@

We will attempt to address it by organizing sessions at conferences
that address the career problem, as well as writing editorials that
can be published in the magazines of our professional societies. These
editorials ought to outline best practices for letter writers that
specifically (i) make clear the contribution of a candidate to their
interdisciplinary area, (ii) the relevance to their home discipline, and
(iii) why writing software is good for the discipline itself.

@@

We have identified a few next steps, even if they may not satisfy the
exact criteria of the SMART system. Specifically:
\begin{itemize}
  \item Find and review the documents listed above, send out summary
    (Wolfgang Bangerth)
  \item Gather feedback from the other members of the group
  \item Write editorial for SIAM (Wolfgang)
  \item Write two emails to each other group members listed above about
    also writing editorials in their communities.
\end{itemize}
In the longer term we hope to gather others in our home disciplines to
let professional organizations weigh in.

>>>>

This group believes that the authors of scientific software provide important
services to departments that are no less than strictly disciplinary
research. Consequently, in order to level the playing field with more
disciplinary candidates for hires, tenure, and promotion requires that
letter writers be aware of how their letters will be read by
committees. It also requires that committees be aware that such
letters often look different and may provide a different perspective
of how a candidate's achievements should be assessed. For example, in
mathematics a typical candidate would be evaluated on the difficulty
and depth of the statements she may have proven in their papers. On
the other hand, an author of mathematical software would likely be
evaluated based on the impact of her software, or the number of citations of
the publications that describe it. She may also be evaluated by how
widely the software is used \textit{outside} mathematics, a criterion
that is rarely used for more disciplinary mathematicians.

@@

Affecting letter writers and readers essentially requires raising
awareness of the issue. Various organizations have attempted to do so
through letters to the community, adjusting definitions of what counts
in science, etc. Specifically, the group is aware of the following
resources:

@@

The group will attempt to address it by organizing sessions at conferences
that address the career problem, as well as writing editorials that
can be published in the magazines of professional societies. These
editorials ought to outline best practices for letter writers that
specifically (i) make clear the contribution of a candidate to their
interdisciplinary area, (ii) the relevance to their home discipline, and
(iii) why writing software is good for the discipline itself.

@@

The group has identified a few next steps, even if they may not satisfy the
exact criteria of the SMART system. Specifically:
\begin{itemize}
  \item Find and review the documents listed above, send out summary
    (Wolfgang Bangerth)
  \item Gather feedback from the other members of the group
  \item Write editorial for SIAM (Wolfgang)
  \item Write two emails to each other group members listed above about
    also writing editorials in their communities.
\end{itemize}
In the longer term, the group hopes to gather others in the home disciplines of its members to
let professional organizations weigh in.

### 2017-04-28 WSSSPE4_report/wg_open_research_index.tex
tense and person changes

The aim of this group is to investigate the building of an index of research products in an open sustainable manner.  Our goal is not to eliminate commercial products, but to build on what is there and provide data and services that are missing.

@@

Alternatively or additionally, we could build a mailing list for us and discuss further, depending on how receptive others are to this idea.  At WSSSPE4, there seemed to be enough interest to do this, so we set up a Slack channel within the WSSSPE team, called \texttt{\#wg-open-research-idx}. (See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)

@@

\begin{enumerate}
\item Look at curriculum lattes (Brazil) and/or Researchfish (UK).  They have some problems that people do not like, we could check what features they has and what the issues are.

@@

\item What research products should we track?

@@

\item Need agreements with publishers, repositories, services
\begin{itemize}
    \item Some of these will have costs, though some may be free
    \item Need to determine costs
    \item Perhaps publishers would pay us to be listed eventually, as we gain power
    \item To some extent, we would be competing with some of the publishers? products
\end{itemize}

@@

\item Apply for initial funding. For the US, we discussed the Arnold Foundation Open Science award as a planning activity (who, due Dec 15), though without a PI, this seems unlikely to be done.
For the UK, we didn't have a suggestion.

>>>>

The aim of this group is to investigate the building of an index of research products in an open sustainable manner.  Its goal is not to eliminate commercial products, but to build on what is there and provide data and services that are missing.

@@

Alternatively or additionally, the group could build a mailing list for us and discuss further, depending on how receptive others are to this idea.  At WSSSPE4, there seemed to be enough interest to do this, so the group set up a Slack channel within the WSSSPE team, called \texttt{\#wg-open-research-idx}. (See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)

@@

\begin{enumerate}
\item Look at curriculum lattes (Brazil) and/or Researchfish (UK).  They have some problems that people do not like, the group could check what features they have and what the issues are.

@@

\item What research products should the index track?

@@

\item Need agreements with publishers, repositories, services
\begin{itemize}
    \item Some of these will have costs, though some may be free
    \item Need to determine costs
    \item Perhaps publishers would pay us to be listed eventually, as we gain power
    \item To some extent, the group would be competing with some of the publishers' products
\end{itemize}

@@

\item Apply for initial funding. For the US, the group discussed the Arnold Foundation Open Science award as a planning activity (who, due Dec 15), though without a PI, this seems unlikely to be done.
For the UK, there wasn't a specific suggestion.

### 2017-04-28 WSSSPE4_report/wg_prototyping.tex
tense and person changes

Our general plan is to develop a functional (meta-)prototype before WSSSPE5, and
if useful, depart from it towards a larger software infrastructure tied to
particular cyberinfrastructure resources.

>>>>

The group's general plan is to develop a functional (meta-)prototype before WSSSPE5, and
if useful, depart from it towards a larger software infrastructure tied to
particular cyberinfrastructure resources.

### 2017-04-28 WSSSPE4_report/wg_social.tex
updating participants lists in working groups to be consistent
removing more author name notes
tense and person changes

\note{from Stuart}

@@

This working group was motivated by the goal of building better connections between academic researchers who are studying topics in or relating to software sustainability with practitioners, managers, and administrators who are working in the area of software sustainability. In any domain, bringing research and practice closer together a mutually beneficial goal, but also has its challenges. This group met to discuss existing research projects and findings that might be relevant for RSEs and others in the software sustainability domain, then identified several gaps and challenges to tackle in bringing research and practice together.

@@

Stuart Geiger <stuart@stuartgeiger.com>, Lorraine Hwang <ljhwang@ucdavis.edu>, Robert McDonald <rhmcdona@indiana.edu>

@@

%What tasks will the working group undertake
The working group planned to briefly survey existing literature to get a better sense of the academic research landscape, facilitate some initial dialog between academic researchers and practitioners at WSSSPE4, then identify needed actions that would be mutually beneficial to researchers and practitioners. This plan resulted in the creation of the following SMART steps listed in the next section.

>>>>

%\note{from Stuart}

@@

This working group is motivated by the goal of building better connections between academic researchers who are studying topics in or relating to software sustainability with practitioners, managers, and administrators who are working in the area of software sustainability. In any domain, bringing research and practice closer together a mutually beneficial goal, but also has its challenges. This group met to discuss existing research projects and findings that might be relevant for RSEs and others in the software sustainability domain, then identified several gaps and challenges to tackle in bringing research and practice together.

@@

\begin{itemize}
  \item Stuart Geiger <stuart@stuartgeiger.com>
  \item Lorraine Hwang <ljhwang@ucdavis.edu>
  \item Robert McDonald <rhmcdona@indiana.edu>
\end{itemize}

@@

%What tasks will the working group undertake
The working group plans to briefly survey existing literature to get a better sense of the academic research landscape, facilitate some initial dialog between academic researchers and practitioners at WSSSPE4, then identify needed actions that would be mutually beneficial to researchers and practitioners. This plan resulted in the creation of the following SMART steps listed in the next section.

### 2017-04-28 WSSSPE4_report/wg_soft_eng.tex
updating participants lists in working groups to be consistent
some misc cleanup/consistency fixes
tense and person changes

%Introduction to group here, including the overall objective of work in this area.
This working group concerned itself with identifying processes that are not adequately
covered by general software engineering. Verification and testing is
the first subtopic that we address.  
% there is a growing awareness about the lack of trust
% and reproducibility in many computational results. 
%\katznote{this paragraph repeats some of the previous one - they probably should be merged}
Computational science and engineering applications have many moving
parts that need to interoperate with one another. The accuracy and
reliability of results produced by scientific software depends not
only on the individual components behaving correctly, but also on the
validity of their interactions. 
% Therefore, a rigorous verification
% process and a robust testing regime are critical requirements for
% scientific software.  There have been instances where inadequate
% verification has resulted in publication of wrong results and later
% retraction of publications.
As scientific understanding grows,
the corresponding computational software models are refined, leading
to more complex codes. Increasing complexity makes them more prone to
defects, not only in individual code units, but also in interaction
among units. Therefore, a strong verification process combined with a
rigorous testing regime plays a critical role in the prevention of
generating incorrect scientific results. However, most science teams
struggle to find a good solution for themselves. Causes range from
lack of exposure to the practices, to distrust of adopting practices
because they do not meet the needs of the teams developing such
software.  The current focus of our effort is on testing because it is the first step
towards building a software processes that can lead to provenance and
reproducibility, the hallmarks of quality science.

\subsubsection{Participants}
\begin{itemize}
\item Mark Abraham
\item Anshu Dubey 
\item Hans Fangohr
\item Dominic Kempf
\item Eric Seidel
\end{itemize}

@@

\subsubsection{Gap or challenge}
%What is the gap or challenge being addressed?
Computational science code developers often lack of exposure to
regular testing and its benefits. Good developers will test their code to
verify that it operates as expected; however, they may not appreciate
that without regular testing, defects can be introduced
inadvertently. An even bigger challenge is that those who understand
the importance of regular testing do not often find much help from
software engineering literature. There is a significant gap between
the testing gospel and its applicability to computational science. This
gap leads to frustration and abandonment of the good with the bad. Some
relevant literature exists, in particular experiences from
practitioners in computational science who developed their own
solutions. However, this literature is scattered among many different
forums, and can be challenging to find. Our working group aims to
address this gap by curating the existing content and contributing
content where none exists.

@@

%What people, groups, or resources are needed.
The working group will benefit from wide participation by developers
of large computational science codes. The reason is that the
management of such codes becomes intractable without adopting some
software process and a testing regime. The experiences and
customizations vary, and the community will benefit from hearing about
as many as possible. The seed resources required are fairly
minimal. We have started a git repository for collecting the
existing references. That, and a few volunteers reading through the
references is all we need in the beginning. As we gather more
knowledge and pinpoint gaps, we may need more resources to reach out
to a wider group of developers to collect more information.

@@

%What are the first SMART steps proposed?
Our first few SMART steps are : 
\begin{itemize}
\item Create a channel in wssspe.slack.com -- {\em done}
\item Create github repository
  {\url{http://github.com/wssspe/testing-in-science}} -- {\em done}
\item Find and gather existing publications in repository -- {\em ongoing}
\item Review and summarize material. Decide whether we consider this
  sufficient. If yes, then we will put brief report together and
  conclude the working group.
\item If we do not consider the material adequate, we start research
  and gather methodologies for testing in science 
\item Write a document accessible to computational science and
  software engineering community, and publish document at a citeable forum.
\end{itemize}

@@

%How could a reader get more information or get more involved?
Readers interested in getting more information should get in touch
with a member of the working group. The working group
has a channel in the WSSSPE Slack. The channel is called
{\it wg-testing-in-science}.
(See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)
Additionally, a git
repository (\url{https://github.com/WSSSPE/WG-Best-Practices.git}) exists for contributing content and reference to, and
curation of the existing literature on this topic.

>>>>

%Introduction to group here, including the overall objective of work in this area.
This working group is concerned with identifying processes that are not adequately
covered by general software engineering. Verification and testing is
the first subtopic that it addresses.
% there is a growing awareness about the lack of trust
% and reproducibility in many computational results.
%\katznote{this paragraph repeats some of the previous one - they probably should be merged}
Computational science and engineering applications have many moving
parts that need to interoperate with one another. The accuracy and
reliability of results produced by scientific software depends not
only on the individual components behaving correctly, but also on the
validity of their interactions.
% Therefore, a rigorous verification
% process and a robust testing regime are critical requirements for
% scientific software.  There have been instances where inadequate
% verification has resulted in publication of wrong results and later
% retraction of publications.
As scientific understanding grows,
the corresponding computational software models are refined, leading
to more complex codes. Increasing complexity makes them more prone to
defects, not only in individual code units, but also in interaction
among units. Therefore, a strong verification process combined with a
rigorous testing regime plays a critical role in the prevention of
generating incorrect scientific results. However, most science teams
struggle to find a good solution for themselves. Causes range from
lack of exposure to the practices, to distrust of adopting practices
because they do not meet the needs of the teams developing such
software.  The current focus of our effort is on testing because it is the first step
towards building a software processes that can lead to provenance and
reproducibility, the hallmarks of quality science.

\subsubsection{Participants}
\begin{itemize}
\item Mark Abraham <mjab@kth.se>
\item Anshu Dubey <adubey@anl.gov>
\item Hans Fangohr <fangohr@soton.ac.uk>
\item Dominic Kempf <dominic.kempf@iwr.uni-heidelberg.de>
\item Eric Seidel <eseidel@cs.ucsd.edu>
\end{itemize}

@@

\subsubsection{Gap or challenge}
%What is the gap or challenge being addressed?
Computational science code developers often lack of exposure to
regular testing and its benefits. Good developers will test their code to
verify that it operates as expected; however, they may not appreciate
that without regular testing, defects can be introduced
inadvertently. An even bigger challenge is that those who understand
the importance of regular testing do not often find much help from
software engineering literature. There is a significant gap between
the testing gospel and its applicability to computational science. This
gap leads to frustration and abandonment of the good with the bad. Some
relevant literature exists, in particular experiences from
practitioners in computational science who developed their own
solutions. However, this literature is scattered among many different
forums, and can be challenging to find. This working group aims to
address this gap by curating the existing content and contributing
content where none exists.

@@

%What people, groups, or resources are needed.
The working group will benefit from wide participation by developers
of large computational science codes. The reason is that the
management of such codes becomes intractable without adopting some
software process and a testing regime. The experiences and
customizations vary, and the community will benefit from hearing about
as many as possible. The seed resources required are fairly
minimal. The group has started a git repository for collecting the
existing references. That, and a few volunteers reading through the
references is all the group needs in the beginning. As it gathers more
knowledge and pinpoint gaps, it may need more resources to reach out
to a wider group of developers to collect more information.

@@

%What are the first SMART steps proposed?
Our first few SMART steps are :
\begin{itemize}
\item Create a channel in wssspe.slack.com -- {\em done}
\item Create github repository
  {\url{http://github.com/wssspe/testing-in-science}} -- {\em done}
\item Find and gather existing publications in repository -- {\em ongoing}
\item Review and summarize material. Decide whether we consider this
  sufficient. If yes, then the group will put a brief report together and
  conclude the working group.
\item If the group does not consider the material adequate, it will start research
  and gather methodologies for testing in science
\item Write a document accessible to computational science and
  software engineering community, and publish document at a citable forum.
\end{itemize}

@@

%How could a reader get more information or get more involved?
Readers interested in getting more information should get in touch
with a member of the working group. The working group
has a channel in the WSSSPE Slack. The channel is called
\texttt{\#wg-testing-in-science}.
(See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)
Additionally, a git
repository (\url{https://github.com/WSSSPE/WG-Best-Practices.git}) exists for contributing content and reference to, and
curation of the existing literature on this topic.
